### 2018-07-09 latex/results.tex
results edited except last subsection

\subsection{Functional properties of NamPRT and NNMT have evolved to maximize NAD-dependent signaling}

@@

To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we have limited information about expression levels of enzymes or changes of kinetic constants during evolution, we initially used the kinetic constants found for human or yeast enzymes for all systems analysed and used equal amounts of enzymes for all reactions\todo{Shorten or split sentence and correct "equal amounts"}. Wherever possible, we did not only include substrate affinities but also known product inhibitions or inhibition by downstream metabolites. As we assume that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.

First, we addressed the unexpected correlation between the selection for co-occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families. We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT. As shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A) while reducing the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B). The decline in NAD concentration can be compensated by a higher expression of NamPRT, further increasing NAD consumption flux (dashed lines in fig.~\ref{fig:NNMT_NAD_flux}A and~B).

These results can be explained when looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirtuin~1. Most NAD-consuming enzymes are inhibited by their product Nam, explaining why the presence of NNMT enables higher NAD consumption fluxes. At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is lower than in the system without NNMT.

As measured kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008}, we analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux. In the absence of NNMT, a change in $K_{M}$ has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B). In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values (fig.~\ref{fig:NamPRT_affinity_Nam}C and~D).

Remarkably, NAD concentration and consumption flux are both considerably affected by cell division rates in a system without NNMT, at least if the enzyme expression is kept constant at different cell division rates. Of course, this is an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead. However, there seems to be a tradeoff between maintainable NAD concentration and consumption flux in the absence of NNMT. In contrast, in the presence of NNMT, NAD consumption rates and concentrations are almost independent of cell division rates.

Figure~\ref{fig:NamPRT_affinity_Nam}E and~F shows a direct comparison of different affinities of NamPRT for Nam with and without NNMT. At an affinity of $K_{M}$ = 1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is higher with NNMT than without, except for low cell division rates (fig.~\ref{fig:NamPRT_affinity_Nam}E). At a lower affinity of 5\,nM, increase in consumption rate even more pronounced. The NAD concentration is lower with NNMT than without for both tested $K_{M}$ values (fig.~\ref{fig:NamPRT_affinity_Nam}F).

As pathway dynamics are not solely dependent on one enzyme, we wanted to estimate the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate. In figure~\ref{fig:optimal_substr_affinities} is shown that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes. Thus, a further increase of the affinity of NamPRT for Nam or a change in the affinity of NNMT for Nam would not provide any advantage.

@@

As our simulations suggested that NNMT might have exerted an evolutionary pressure on the development of NamPRT, we created a multiple sequence alignment to see if we can find sequence variations in NamPRT upon the appearance of NNMT. The multiple sequence alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1. We found that most Deuterostomes that possess only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme. This insert overlaps with a predicted weak nuclear localisation signal (NLS), that is lost when the insert is removed. The sequence corresponds to a loop at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g. structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}). The loop is connected to one of the $\beta$-sheets involved in substrate binding.

From these observations, we derived two possible hypotheses regarding the role of the loop in NamPRT function. The first hypothesis was that the presence of the loop changes the subcellular localisation of NamPRT, since the loop is part of an NLS. To test this hypothesis, we recombinatly expressed wildtype NamPRT and a mutated version lacking amino acids 42 to 51. Both, wildtype and mutant, were modified to carry a FLAG-tag for easier visualisation. Immunofluorescence imaging showed a mixed cytosolic nuclear localisation for the wildtype and the mutant NamPRT (fig.~\ref{fig:unresolved_loop}C). The presence of the loop does not seem to have any effect on the subcellular localisation.

The second hypothesis was that the sequence insertion influences the affinity of NamPRT, which would be in line with the predictions from our simulations. We recombinantly expressed both wildtype and mutant NamPRT in \textit{E.~coli}. After purification (suppl. fig.~S3), we measured the activity using NMR detection of NMN in the presence and absence of ATP. The enzymatic activity of the mutant enzyme lacking the loop was much lower than the activity of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D). An increase of the substrate concentration had a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity. In contrast to wildtype NamPRT, the mutant does not exhibit an increased activity in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).

>>>>

\subsection{Functional properties of NamPRT and NNMT have evolved to maximize NAD-dependent signalling}

@@

To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we have limited information about expression levels of enzymes, we initially used equal amounts for all enzymes. As we have very few cross species kinetic data, we were furthermore mainly relying on kinetic constants found for human or yeast enzymes.  Wherever possible, we did not only include substrate affinities but also known product inhibitions or inhibition by downstream metabolites. As we in addition assumed that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.

First, we addressed the unexpected correlation between the selection for co-occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families. We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT. To achieve free NAD concentration in the range reported in the literature and due to the very low turnover of NamPRT, we had to use 10 fold higher NamPRT levels compared to other enzymes and adjusted the amount of NMNAT to the same level to avoid that the NAD synthesis rates are limited by this enzyme.  As shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A) while reducing the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B). The decline in NAD concentration can be compensated by a higher expression of NamPRT, further increasing NAD consumption flux (dashed lines in fig.~\ref{fig:NNMT_NAD_flux}A and~B).

These results can be explained by looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirtuin~1. Most NAD-consuming enzymes are inhibited by their product Nam. Thus the presence of NNMT enables higher NAD consumption fluxes, by removing excess Nam from the cells. At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is of course lower than in the system without NNMT.

As measured kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008}, we analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux. In the absence of NNMT, a variation of the substrate affinity of NamPRT for Nam, has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B). In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values (fig.~\ref{fig:NamPRT_affinity_Nam}C and~D).

Remarkably, NAD concentration and consumption flux are both considerably affected by cell division rates in a system without NNMT, at least if the enzyme expression is kept constant at different cell division rates. Of course, this is an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead. Nevertheless, in the absence of NNMT there seems to be a tradeoff between maintainable NAD concentration and consumption flux. In contrast, in the presence of NNMT, NAD consumption rates and concentrations are almost independent of cell division rates.

Figure~\ref{fig:NamPRT_affinity_Nam}E and~F show a direct comparison of different affinities of NamPRT for Nam with and without NNMT. At an affinity of $K_{M}$ = 1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is only higher with NNMT than without at low cell division rates (fig.~\ref{fig:NamPRT_affinity_Nam}E). If the affinity of NamPRT is high enough, consumption rates are always higher with NNMT than without it. The NAD concentration is always lower with NNMT (fig.~\ref{fig:NamPRT_affinity_Nam}F).

As pathway dynamics are not solely dependent on one enzyme, we wanted to estimate the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate. In figure~\ref{fig:optimal_substr_affinities} is shown that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes. Thus, a further increase of the affinity of NamPRT for Nam or a change in the affinity of NNMT for Nam would not provide any further  advantage.

@@

As our simulations suggested that NNMT might have exerted an evolutionary pressure on the development of NamPRT, we created a multiple sequence alignment to see if we can find sequence variations in NamPRT upon the appearance of NNMT. The multiple sequence alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1. We found that most Deuterostomes that possess only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme. This insert overlaps with a predicted weak nuclear localisation signal (NLS), that is lost when the insert is removed. The sequence corresponds to a loop at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g. structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}). The loop is connected to one of the $\beta$-sheets involved in substrate binding \cite{PMID:19666527}.

From these observations, we derived two possible hypotheses regarding the role of the loop in NamPRT function. The first hypothesis was that the presence of the loop changes the subcellular localisation of NamPRT, as it is overlapping with a predicted NLS. To test this hypothesis, we recombinatly expressed FLAG-tagged  wildtype and mutant NamPRT in Hela S3 cells. Immunofluorescence imaging showed a mixed cytosolic nuclear localisation for the wildtype and the mutant NamPRT (fig.~\ref{fig:unresolved_loop}C). Thus, the presence of the loop does not compromise the nuclear localisation.

The second hypothesis was that the sequence insertion influences the affinity of NamPRT, which would be in line with the predictions from our simulations. We recombinantly expressed both wildtype and mutant NamPRT in \textit{E.~coli}. After purification (suppl. fig.~S3), we measured the activity using NMR detection of NMN in the presence and absence of ATP. The enzymatic activity of the mutant enzyme lacking the loop was much lower than the activity of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D). Furthermore, an increase of the substrate concentration had a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity. In contrast to wildtype NamPRT, the mutant does not exhibit an increased activity in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).
